Emergent dual scaling of riverine biodiversity

### 2021-03-23 coverletter.Rmd
Update coverletter.Rmd

We are excited to submit our manuscript entitled 'Ecosystem size and complexity dictate riverine biodiversity' to PNAS. Given the unprecedented rate of global biodiversity loss, a crucial task for ecologists is to guide conservation efforts by identifying key environmental factors that regulate biodiversity. A prevailing wisdom suggests that species diversity is primarily demined by area; larger ecosystems support more species (i.e., the Species Area Relationship; SAR). This ecological pattern has been found in many taxa and geographic regions, forming the foundation of current conservation schemes. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas, which encompass more than 30 million km2 globally. However, what is largely overlooked in current biodiversity research is the role of a 'scale-invariant' complexity of ecosystem's geometric structure (hereafter, 'ecosystem complexity'). Since ecosystem complexity is independent of spatial scale or ecosystem size, existing theories are inappropriate to predict biodiversity patterns in spatially complex ecosystems.

Here, we offer a synthetic framework integrating ecosystem size and complexity influences on biodiversity patterns in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few. In this widespread yet largely overlooked fractal ecosystem's geometry, branching patterns remain constant across a wide range of spatial scales. Even so, the complexity of branching patterns forms a physical template providing a wide spectrum of environmental niches for living organisms. In this research, using theory and extensive fish community data from two distinct geographic regions, **we found that ecosystem size and complexity play comparable roles in regulating riverine biodiversity; ecosystem size alone was insufficient to predict biodiversity patterns.** Larger and more branched 'complex' river networks supported greater species richness due to increased space and habitat heterogeneity. In particular, the effect of ecosystem complexity was striking in its magnitude. A unit increase of ecosystem complexity (the prevalence of branching) resulted in an increase of species diversity that would require 1911 km2 in the areal area. Further, these patterns have consistently emerged regardless of ecological traits of species and geographic regions, suggesting that the dual control of riverine biodiversity may be pervasive across taxonomic groups and geographic regions.

Our research will attract a broad audience of PNAS for the following reasons. First, our proposed framework may apply beyond branching ecosystems as both ecosystem size and complexity represent a physical template of the ecosystem's environmental heterogeneity. Second, the results of the present study provide the ecological foundation for complexity-based ecosystem management - an important perspective overlooked in the current schemes of biodiversity conservation. Hence, we have confidence that our research not only provides fundamental insights into regulating factors of biodiversity but also has far-reaching implications for biodiversity conservation.

>>>>

We are excited to submit our manuscript entitled 'Emergent dual scaling of riverine biodiversity' to *Proceedings of the National Academy Sciences*.

In the Anthropocene, it is more imperative than ever to identify ecological factors that underpin diverse ecological communities in nature. Much attention has been paid to the ubiquitous power-law scaling of species richness and area, in which species richness increases with area in a decelerating way. Ecologists have used this ecological *law* to inform many conservation strategies, including marine and terrestrial protected areas that currently encompass > 30 million km^2^ globally. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because 'scale-invariant' complexity of ecosystem structure may drive ecological dynamics in space. Branching river networks are an outstanding example of such complex ecosystems, where geometric ecosystem structure (e.g., branching patterns) remains constant across a wide range of spatial scales.

Here, we used theory and analysis of large fish community data to show that **ecosystem complexity of branching networks (measured as the prevalence of branching) plays a crucial role in scaling metacommunity-level biodiversity in rivers (**$\gamma$ **diversity); ecosystem size and other macro-scale drivers (e.g., climates and land use patterns) alone were insufficient to describe biodiversity patterns in complex river networks.** Our theory predicted that, regardless of simulated species' traits, more branched 'complex' river networks supported greater species richness due to increased habitat heterogeneity. This relationship had a characteristic of power-law, revealing the overlooked scaling relationship in complex branching networks. Importantly, the complexity effect was comparable to the size effect - hence, riverine biodiversity should follow a dual scaling law along the two orthogonal dimensions of ecosystem structure, i.e., size and scale-invariant complexity.

We confirmed the theoretical predictions using extensive fish community data, which included 180 replicates of metacommunities (i.e., watersheds) across two geographic regions (Japan and the US). The dual scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fish fauna with distinct evolutionary histories. This result agrees elegantly with theoretical predictions, in which species' traits played a minor role in shaping the dual power-law scaling. In particular, the effect of ecosystem complexity was striking in its magnitude. A unit increase of ecosystem complexity (the prevalence of branching) resulted in an increase of species diversity that would require 1911 km^2^ in the areal area.

Our study distinguishes itself from other studies in three ways. First, we offer a synthetic framework integrating ecosystem size and complexity influences on biodiversity, which should apply beyond branching ecosystems. Second, unlike most previous studies focused on local biodiversity patterns ($\alpha$ diversity), our empirical analysis used metacommunities as replicates of statistical analysis. Therefore, this study provides crucial insights into how ecological communities are structured *at the metacommunity level*. Lastly, the present study provides the foundation for complexity-based ecosystem management - an important perspective overlooked in current conservation schemes. Hence, we have confidence that our research provides fundamental insights into regulating factors of biodiversity and has far-reaching implications for biodiversity conservation.
